---
title: "Introduction to Phylogenetics in R"
output:
  html_document:
    df_print: paged
---

# An introduction to phylogenetic comparative methods in R
This tutorial is intended to get you familiar with the R environment for conducting
phylogenetic comparative analyses.

## I. Preparing the R environment
```{r echo=FALSE}
options(max.print=50)
```

### Installing packages
To install a package directly from CRAN, use:
```{r eval=FALSE}
install.packages("geiger")
install.packages("phytools")
install.packages("treeplyr")
install.packages("rotl")
install.packages("phylolm")
install.packages("datelife")
```

Load the package using 'library'.
```{r}
library(geiger)
library(phytools)
library(treeplyr)
library(rotl)
library(datelife)
library(phylolm)
```

For phylogenetic comparative methods, there are a number of packages available. One place
to find them all is the 'Phylogenetics Task View' run by Brian O'meara: (http://cran.r-project.org/web/views/Phylogenetics.html)
All of these packages can be installed at once using the following lines (however, please note that
this process can take a considerable amount of time!)

```{r eval=FALSE}
install.packages("ctv")
install.views("Phylogenetics")
```

### II. Reading in a tree & tree data structures
There are a number of functions to read in phylogenetic tree data into R. We are going
to use as our example tree phylogeny of Muroid rodents (Schenk, Rowe & Steppan, 2013; Systematic Biology).
The tree can be downloaded from my github site, or it be accessed directly from treebase (submission 12303).

```{r}
tree <- read.tree("../data/schenk.tre")
tree
```

We can plot the tree:
```{r}
plot(tree)
plot(tree, type="fan", cex=0.25)
```

You may notice that the help file for `plot` is not very helpful for options for phylogenies: 
```{r}
?plot
```

This is because `plot` is a "method" that behaves differently on objects of different classes. 
```{r}
plot
methods(plot)
class(tree)
?plot.phylo
```

How is a tree stored? How can it be manipulated and modified? 
A tree object in R is a data structure called a "list", and given a special class, called a "phylo" object.
It will be worth your time getting familiar with [data structures in R](http://adv-r.had.co.nz/Data-structures.html).

`str` is a useful commands for determining what is in an R object:
```{r}
str(tree)
```

Each element of the list can be accessed by multiple means:
```{r}
tree['tip.label']
tree[['tip.label']]
tree[[3]]
tree$tip.label
```

The structure of the tree is stored in the edge matrix:
```{r}
tree$edge
```

And branch lengths are stored in the list element `edge.length`
```{r}
tree$edge.length
```

Tips can be dropped with the function `drop.tip`:
```{r}
#tree before:
tree
#tree after dropping two species:
drop.tip(tree, c("Eliomys_quercinus", "Sicista_tianshanica"))
```

### III. Matching a tree with a dataset

```{r}
pantheria <- read.csv("../data/pantheria.csv")
rownames(pantheria) <- gsub(" ", "_", pantheria$MSW05_Binomial)
head(pantheria)
```

```{r}
td <- make.treedata(tree, pantheria)
td
```

```{r}
attributes(td)$dropped
```

```{r}
colnames(td$dat)
```

```{r}
td2 <- mutate(td, lnMass = log(X5.1_AdultBodyMass_g),lnBMR = log(X18.1_BasalMetRate_mLO2hr), desert=ifelse(X28.1_Precip_Mean_mm < 21, 1, 0)) %>% filter(., !is.na(lnMass), !is.na(lnBMR)) %>% select(., lnMass, lnBMR, desert)
td2
```

```{r}
plot(td2[['lnMass']], td2[['lnBMR']], pch=21, bg=td2[['desert']]+2, xlab="lnMass", ylab="lnBMR")
```

## IV. Visualization
We can visualize the distribution of the data on the tree using a "traitgram":
```{r}
phenogram(td2$phy, td2[['lnMass']], spread.labels=FALSE)
```

We can also calculate the node contrasts (standardized differences) to see where the big changes are on the tree:
```{r}
picMass <- pic(td2[['lnMass']], td2$phy)
plot(tree, cex=0.5)
nodelabels(pch=21, bg=topo.colors(100)[round((picMass-min(picMass))/diff(range(picMass))*100,0)])
```

Phytools' scattergram can be useful for exploring correlations between traits across the phylogeny:
```{r}

fancyDat <- as.matrix(td2$dat[1:3])
rownames(fancyDat) <- td2$phy$tip.label
fancyTree(td2$phy, type="scattergram", X=fancyDat, fsize=0.3)

```

## V. Simple analyses of traits

### Discrete trait models
We generally use "continuous-time Markov models" to model discrete traits on the phylogeny. These models are
used to model transitions between discrete character states, whether they are the presence/absence of phenotypic trait or nucleotide in a sequence of DNA. We can fit such a model to a trait and a phylogeny and then draw a possible reconstructed history of the trait under the model, which is called ancestral state reconstruction. Here we make "stochastic character maps", which are draws from the posterior distribution of ancestral state reconstructions under the model. This illustrates how uncertain ancestral state reconstruction is!

```{r}
tdDiscrete <- filter(td2, !is.na(desert)) %>% mutate(., desert2=desert+1)
simmap <- make.simmap(tdDiscrete$phy, tdDiscrete[['desert2']], model="ER", nsim=9)
par(mfrow=c(3,3), mar=c(0,0,0,0))
for(i in 1:9){
  plot(simmap[[i]], ftype="off")
}
```

We assumed in the previous model that transitions in and out of desert habitats happened at equal frequency. If we relax this assumption and allow these to be asymmetric, we can see that rodents move out of desert habitats at 3 times the rate they move in. This will change the reconstructed histories!

```{r}
simmap <- make.simmap(tdDiscrete$phy, tdDiscrete[['desert2']], model="ARD", nsim=9)
par(mfrow=c(3,3), mar=c(0,0,0,0))
for(i in 1:9){
  plot(simmap[[i]], ftype="off")
}
```

### Continuous trait models
For continuous traits, we generally use "Gaussian models" that result in multivariate normal distributions.
The most basic and familiar of these is Brownian motion. We can fit a Brownian motion model using the function `fitContinuous` from the `geiger` package. 

```{r}
mBM <- fitContinuous(td2$phy, td2[['lnMass']], model="BM")
mBM
```

An more general model than the Brownian motion model is the Ornstein-Uhlenbeck model, which has one additional parameter describing the central tendency of traits to evolve toward an intermediate optimum (think of a rubber band pulling back towards an optimum trait value).
```{r}
mOU <- fitContinuous(td2$phy, td2[['lnMass']], model="OU")
mBM
```

Another model is one in which the rate of evolution is initially rapid, but slows over time. This is called
the early-burst model (EB). Let's fit all models simultaneously using a loop. This time, we'll use the BMR data instead of the mass data:
```{r}
models <- c("BM", "OU", "EB")
mFits <- list()
for(i in 1:length(models)){
  mFits[[i]] <- fitContinuous(td2$phy, td2[['lnBMR']], model=models[i])
}
```

Model selection can be conducted by comparing AIC values:
```{r}
aiccs <- lapply(mFits, function(x) x$opt$aicc)
names(aiccs) <- models
aiccs
```

Which model did best? How do we interpret this? 

## VI. Phylogenetic Regression
When we want to relate two traits to each other, we can make use of "Phylogenetic Generalized Least Squares Regression" or PGLS. This is equivalent to Independent Contrasts with Brownian Motion. The best package for using various types of this approach is `phylolm` written by Lam Ho, Cecile Ane, and others. 

```{r}
rownames(td2$dat) <- td2$phy$tip.label
ols.lm1 <- lm(lnBMR ~ lnMass, data=td2$dat)
pgls.lm1 <- phylolm(lnBMR ~ lnMass, data=td2$dat, phy=td2$phy, model="BM")
summary(ols.lm1)
summary(pgls.lm1)
```

The results aren't that different, in this case! Try a phylogenetic model with an interaction between desert and lnMass. 

```{r}

```

Check out the other models you can fit:
```{r}
?phylolm
```

These all have various assumptions, one in particular, the `lambda` transform, allows you to slide between an OLS and PGLS regression depending on the 
"Phylogenetic signal" measured in your data. 
```{r}
lambda.lm <- phylolm(lnBMR ~ lnMass, data=td2$dat, phy=td2$phy, model="lambda")
summary(lambda.lm)
```

Let's plot the results: 

```{r}
plot(td2[['lnMass']], td2[['lnBMR']], xlab="lnMass", ylab="lnBMR", pch=21, bg=td2[['desert']]+1)
abline(ols.lm1, col="red")
abline(pgls.lm1, col="blue")
abline(lambda.lm, col="purple")
```

## VII. Getting a phylogeny from a list of taxa

Most comparative analyses require a time-calibrated phylogeny. The gold standard is to estimate this phylogeny from molecular sequence data, and time-calibrate the resulting phylogeny with fossil calibrations. However, some resources exist for drawing on existing phylogenetic and taxonomic information. One project spearheading the effort to synthesize existing phylogenetic knowledge is the OpenTree of Life (OTL) that has excellent API resources that can be used to obtain synthetic phylogenies, as well as other services such as taxonomic name resolution (this is huge, you will spend a huge amount of time dealing with names). 

```{r}
taxalist <- c("Picoides_borealis", "Carpodacus_mexicana", "Hyla_regilla", "Lithobates_catesbeiana", "Cambarincola", "Cambarus_chasmodactylus", "Pipra_filicauda", "Chrysopelia_paradisi", "Norops_sagrei", "Carollia_perspicillata", "Piper peltatum", "Taeniopygia_guttata", "Plethodon_hubrichti", "Felus_domestica", "Aedes_aegypti", "Plasmodium_vivax")

```

Let's use OpenTree of Life's powerful taxonomic name resolution services, which draws on NCBI, GBIF, Fishbase, and others to build a comprehensive taxonomy for all life.
```{r}
ottTable <- tnrs_match_names(taxalist)
ottTable

```

Now we can use the Open Tree Taxonomy (OTT) IDs to obtain a synthetic tree of only those taxa.
```{r}
synth_tree <- rotl::tol_induced_subtree(ott_ids=ottTable$ott_id)
```

```{r}
plot(synth_tree, cex=0.75)
```


